Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-29 parallel-tractability-J-v9/parallel-tractability-J.tex
added the analysis of complexity of Algorithm A_{opt}

\subsection{A Basic Parallel Algorithm}

@@

\noindent\texttt{Algorithm~$\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,\footnote{Suppose that each processor
    can use $O(1)$ time units to access the state of ground atoms, i.e., whether this ground
    atom has been added to the materialization graph. This can be implemented by maintaining an
    index of polynomial size.}
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupies constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{bsc}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus, 
the whole computing time of Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+l\cdot c_2$
where $l$ denotes the number of iterations of \ref{alg1:updateG}.

@@

\noindent\texttt{Algorithm~$\mathsf{A}_{opt}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$, the algorithm
returns a materialization graph $\mathcal{G}$ of $P$. Initially $\mathcal{G}$ is empty. The following steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg3:addFacts}
\item Compute $S_{\textit{\tiny rch}}$ by performing \ref{rch} in Algorithm~$\mathsf{OPT}$; use an NC
    algorithm to compute the transitive closure $S^*_{\textit{\tiny rch}}$ (see \ref{transClos} in Algorithm~$\mathsf{OPT}$);
    update $\mathcal{G}$ by performing \ref{updateG}  in Algorithm~$\mathsf{OPT}$.\label{alg3:updateG}
\item If no node has been added to $\mathcal{G}$ (in \ref{alg3:updateG}), terminate,
    otherwise iterate \ref{alg3:updateG}. \label{alg3:halt}\hfill$\Box$
\end{enumerate}

@@

We obtain an NC variant of $\mathsf{A}_{opt}$ analogously to the process for Algorithm~$\mathsf{A}_{bsc}$.
It can be checked that an iteration of \ref{alg3:updateG} in $\mathsf{A}_{opt}$ costs poly-logarithmic
time, since the main part is computing $S^*_{\textit{\tiny rch}}$ by an NC algorithm.
Thus, if the number of iterations of \ref{alg3:updateG} is upper-bounded by a poly-logarithmical function,
$\mathsf{A}_{opt}$ is an NC algorithm.
Analogously to $\mathsf{A}_{bsc}^{\psi}$, we use $\mathsf{A}_{opt}^{\psi}$ to denote an NC variant.
Specifically, for any datalog program $P$, the number of iterations of \ref{alg3:updateG} in $\mathsf{A}_{opt}$
is bounded by $\psi(|P|)$, where $\psi$ is a poly-logarithmically bounded function.

>>>>

\subsection{A Basic Parallel Algorithm}
\label{sec:alg-bsc}

@@

\noindent\texttt{Algorithm~$\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{bsc}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus, 
the whole computing time of Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+l\cdot c_2$
where $l$ denotes the number of iterations of \ref{alg1:updateG}.

@@

\noindent\texttt{Algorithm~$\mathsf{A}_{opt}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$, the algorithm
returns a materialization graph $\mathcal{G}$ of $P$. 
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg3:addFacts}
\item Compute $S_{\textit{\tiny rch}}$ by performing \ref{rch} in Algorithm~$\mathsf{OPT}$; use an NC
    algorithm to compute the transitive closure $S^*_{\textit{\tiny rch}}$ (see \ref{transClos} in Algorithm~$\mathsf{OPT}$);
    update $\mathcal{G}$ by performing \ref{updateG}  in Algorithm~$\mathsf{OPT}$.\label{alg3:updateG}
\item If no node has been added to $\mathcal{G}$ (in \ref{alg3:updateG}), terminate,
    otherwise iterate \ref{alg3:updateG}. \label{alg3:halt}\hfill$\Box$
\end{enumerate}

@@

The analysis of the parallel complexity of Algorithm~$\mathsf{A}_{opt}$ is given as follows.
We use the same symbols $e$, $v$ and $r$ that are used for analyzing the complexity of Algorithm~$\mathsf{A}_{bsc}$
(see Section~\ref{sec:alg-bsc}). Similarly to the analysis for Algorithm~$\mathsf{A}_{bsc}$, it can be checked that 
the number of processors is polynomial of $|\textbf{I}|$, i.e., $r(|\textbf{I}|e)^v$.
We now consider the computing time of Algorithm~$\mathsf{A}_{opt}$. It is obvious that \ref{alg3:addFacts}
\ref{alg3:halt} occupy constant time (denoted by $c_1$). In \ref{alg3:updateG}, the phases of
computing $S_{\textit{\tiny rch}}$ and updating $\mathcal{G}$ also take constant time (denoted by $c_2$)
under the parallel assumption. Recall that the number of constants is at most $|\textbf{I}|e$ (see Section~\ref{sec:alg-bsc}).
Let $w$ and $p$ represent the maximum arity of any predicate and the number of predicates respectively.
We have that the number of all possible substitutions of atoms is at most $p(|\textbf{I}|e)^w$.
The relation $S_{\textit{\tiny rch}}$ consists of pairs of the atoms. Thus, the size of $S_{\textit{\tiny rch}}$
is bounded by $p^2(|\textbf{I}|e)^{2w}$. It can be checked that the computing time of the NC algorithm for computing the transitive
closure of $S_{\textit{\tiny rch}}$ is poly-logarithmical in the size of $\textbf{I}$, more precisely $2\log(p)+2w\log(|\textbf{I}|e)$.
We now have that the total computing time of Algorithm~$\mathsf{A}_{opt}$ is $c_1+lc_3+2lw\log(|\textbf{I}|e)$
where $c_3=c_2+2\log(p)$ and $l$ denotes the number of iterations of \ref{alg3:updateG}.

Algorithm~$\mathsf{A}_{opt}$ can be restricted to an NC algorithm analogously to the process 
for Algorithm~$\mathsf{A}_{bsc}$. That is the number of iterations of \ref{alg3:updateG}
is poly-logarithmically bounded, i.e., it is bounded by $\psi(|\textbf{I}|)$
where $\psi$ is a poly-logarithmical function. In this way, the computing time of 
Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+c_3\psi(|\textbf{I}|)+2w\psi(|\textbf{I}|)\log(|\textbf{I}|e)$,
which is also poly-logarithmical in the size of $\textbf{I}$. We denote the 
NC variant of Algorithm~$\mathsf{A}_{opt}$ by $\mathsf{A}_{opt}^{\psi}$.
